Flabby and injective objects in toposes

### 2018-10-21 paper-flabby-objects.tex
typo

\usepackage[utf8]{inputenc}
\usepackage{amsthm,mathtools,tikz,float,caption,stmaryrd}
\usetikzlibrary{patterns,matrix,decorations.pathreplacing}
\usepackage[all]{xy}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{hyperref}

@@

\newcommand{\xra}[1]{\xrightarrow{#1}}
\newcommand{\XXX}[1]{\textbf{XXX: #1}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Mod}{Mod}
\DeclareMathOperator{\Sh}{Sh}
\DeclareMathOperator{\PSh}{PSh}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\renewcommand{\_}{\mathpunct{.}\,}

@@

\begin{abstract}
  We introduce a general notion of \emph{flabby objects} in elementary toposes
  and study their basic properties. In the special case of localic toposes, this
  notion reduces to the common notion of flabby sheaves, yielding a
  site-independent characterization of flabby sheaves. Continuing a line of
  research started by Roswitha Harting, we use flabby objects to
  show that an internal notion of injective objects coincides with the
  corresponding external notion, in stark contrast with the situation for
  projective objects. As an application, we give internal characterizations of
  sheaf cohomology and show that higher direct images can be understood as
  internal cohomology.
\end{abstract}

@@

Kenney stressed that the notion of an injective set should be regarded as an
interesting strengthening of the constructively rather ill-behaved notion of a
nonempty set~\cite{kenney:injective-choice}. For instance, while the statements
``there is a choice function for every set of nonempty sets'' and even ``there
is a choice function for every set of inhabited sets'' are constructive taboos,
the statement ``there is a choice function for every set of injective sets'' is
constructively neutral. Proposition~\ref{prop:set-of-preimages-flabby} demonstrates
that the notion of a flabby set can be regarded as an interesting intermediate
notion: In the situation of Proposition~\ref{prop:set-of-preimages-flabby}, the
set of preimages is not only not empty or inhabited, but even flabby.

@@

\section{Flabby objects}

@@

If~$R$ is a ring in an elementary topos~$\E$, a similar definition can be given
for~$R$-modules in~$\E$, referring only to the set respectively the object of
linear maps. The condition for an object to be internally injective can be
rephrased in various ways. The following proposition lists five of these
conditions. The equivalence of the first four is due to
Harting~\cite{harting:locally-injective}.

@@

The implication (2)~$\Rightarrow$~(2') employs the
extra left adjoint $p_! : \E/A \to \E$ of $p^* : \E
\to \E/A$~(which maps an object~$(X \to A)$ to~$X$), as in the usual proof that
injective sheaves remain injective when
restricted to smaller open subsets: We have that $p_* \circ [\cdot, p^*(I)]_{\E/A}
\cong [\cdot, I]_\E \circ p_!$, the functor $p_!$ preserves monomorphisms, and one
can check that $p_*$ reflects the property that global elements locally possess
preimages. Details are in~\cite[Thm.~1.1]{harting}.\footnote{Harting formulates
the statement for abelian group objects, and has to assume that~$\E$ contains a
natural numbers object to ensure the existence of an abelian version of~$p_!$.}

@@

Somewhat surprisingly, and in stark contrast with the situation for internally
projective objects (which are defined dually), internal injectivity coincides
with external injectivity for localic toposes.

\begin{thm}\label{thm:injectivity-external-internal}
Let~$I$ be an object of an elementary topos~$\E$. If~$I$ is externally
injective, then~$I$ is also internally injective. The converse holds if~$\E$ is
localic and Zorn's lemma is available.
\end{thm}

@@

The analogue of Theorem~\ref{thm:injectivity-external-internal} for modules
holds as well, if~$\E$ is assumed to have a natural numbers object. The proof
carries over word for word, only referencing
Lemma~\ref{lemma:set-of-extensions-flabby}(2) instead of
Lemma~\ref{lemma:set-of-extensions-flabby}(1).
It seems that Harting was not aware of this, even though she did show that
injectivity of sheaves of modules over topological spaces is a local
notion~\cite{XXX}, as she
(mistakenly) states in~\cite[page~233]{harting:remark} that ``the notions of
injectivity and internal injectivity do not coincide'' for modules.
% XXX comparison with Harting's proof (Barr's lemma...)

@@

\begin{proof}The analogous statements were established purely
intuitionistically in Section~\ref{sect:flabby-sets}, and the stack semantics
is sound with respect to intuitionistic logic.
\end{proof}

@@

\section{Higher direct images as internal sheaf cohomology}

Let~$X$ be a locale and let~$f : Y \to X$ be an over-locale. By the fundamental
relation between locales and topological spaces, this situation arises
for instance, when given a sober topological space and a topological
space over it, as is often the case in algebraic topology or algebraic
geometry. Let a sheaf~$\O_Y$ of rings on~$Y$ be given. Then the
traditional way to define the \emph{higher direct images} of a sheaf~$E$
of~$\O_Y$-modules is to pick an injective resolution~$0 \to E \to I^\bullet$
and set~$R^n f_*(E) \defeq H^n(\Gamma(I^\bullet))$.

@@

Usually one would not expect an internal construction which depends on
arbitary choices to yield a globally-defined sheaf on~$X$ -- following the
definition of the stack semantics we only obtain a family of sheaves defined on
members of some open covering of~$X$; but we verify in
Theorem~\ref{thm:higher-direct-images-as-internal-sheaf-cohomology} that in our
case, it does, and that the resulting sheaf coincides with the~$n$-th higher
direct image of~$E$.

@@

The internal characterization provided by
Theorem~\ref{thm:higher-direct-images-as-internal-sheaf-cohomology} gives, as a
simple application, a geometric explanation that higher direct images along
the identity~$\mathrm{id} : X \to X$ vanish: From the internal point of view
of~$\Sh(X)$, the over-locale~$X$ corresponds to the one-point locale, and the
higher cohomology of the one-point locale vanishes.

@@

\section{Flabby objects in the effective topos}
\label{sect:in-eff}

The notion of flabby objects originates from the notion of flabby shaves and is
therefore closely connected to Grothendieck toposes. Hence it is instructive
to study flabby objects in elementary toposes which are not Grothendieck
toposes, away from their original conceptual home. We begin this study with
establishing the following observations on flabby objects in the effective
topos. We follow the terminology of~\cite{hyland:effective-topos}.

\begin{prop}\label{prop:flabby-effective-sets}
Let~$X$ be a flabby object in the effective topos. Let~$f : X \to X$
be a morphism. If~$X$ is effective, the statement ``$f$ has a fixed point''
holds in the effective topos.
\end{prop}

\begin{prop}\label{prop:semienough-flabby-modules}
Assuming the law of excluded middle in the metatheory, any~$\neg\neg$-separated
module in the effective topos embeds into a flabby module.
\end{prop}

@@

The intuitive
reason for why Proposition~\ref{prop:flabby-effective-sets} holds is the
following. Let~$X$ be a flabby object in the effective topos. Then there is a
procedure which computes for any subterminal~$K \subseteq X$ an element~$x_K$
such that~$K \subseteq \{ x_K \}$. This element might not depend extensionally
on~$K$, but this fine point is not important for this discussion. Let
now~$f : X \to X$ be a morphism. We construct the self-referential subset~$K
\defeq \{ f(x_K) \}$; the formal proof below will indicate how this can be
done. Then~$K \subseteq \{ x_K \}$, so~$f(x_K) = x_K$.

A corollary of Proposition~\ref{prop:flabby-effective-sets} is that the trivial
module is the only flabby module in the effective topos whose underlying
unstructured object is an effective set: Given such a flabby module~$M$, let~$v
\in M$ be an arbitrary element. Then the morphism~$x \mapsto v + x$ has a fixed
point; thus~$v + x = x$ for some element~$x$, and hence~$v = 0$.

It is the self-referentiality which makes the proof of
Proposition~\ref{prop:flabby-effective-sets} work, but the blame for paucity
of flabby objects in the effective topos is to put on the realizers for
statements of the form~``$K = K$'', where~$({=})$ is the nonstandard equality
predicate of the powerobject~$\P(X)$. A procedure witnessing flabbiness has to
compute a reflexivity realizer for a suitable element~$x_K$ from a reflexivity
realizer for a given element~$K$. However, such realizers are not very
informative. Metaphorically speaking, a procedure witnessing flabbiness has to
conjure elements out of thin air.

This problem does not manifest with objects~$X$ which are not effective sets.
Reflexivity realizers for these objects are themselves not very informative;
a procedure witnessing flabbiness therefore only has to turn one kind of
non-informative realizers into another kind. The flabby modules featuring in
the proof of Proposition~\ref{prop:semienough-flabby-modules} will accordingly
not be effective sets.

\begin{proof}[Proof of Proposition~\ref{prop:flabby-effective-sets}]
For any Turing machine~$e$, let~$v_e : |X| \to \Sigma$ be the nonstandard
predicate given by
\begin{multline*}
  v_e(x) = \{ m \in \NN \,|\,
  \text{there is an element~$x_0 \in |X|$ such that} \\
  \text{$e$ terminates with an element of~$\llbracket x_0 = x_0 \rrbracket$ and
  $m \in \llbracket x = x_0 \rrbracket$} \}
\end{multline*}
and let~$K_e \in \Sigma^{|X| \times \Sigma}$ be the nonstandard predicate given by
\[ K_e(x,u) = \llbracket (x = x) \wedge (u \leftrightarrow v_e(x)) \rrbracket. \]
One can explicitly construct a realizer~$a_e$ of the statement~``$K_e = K_e$'',
where~$({=})$ is the nonstandard equality predicate of the object~$\P_{\leq1}(X)$
of subterminals of~$X$. This is where the assumption that~$X$ is effective is
important; without it, we could only verify~``$K_e = K_e$'' where~$({=})$ is
the nonstandard equality predicate of the full powerobject~$\P(X)$.

@@

\begin{proof}[Proof of Proposition~\ref{prop:semienough-flabby-modules}]
Let~$(\Gamma \dashv \Delta) : \Set \to \Eff$ be the inclusion of the
double-negation sheaves. For a~$\neg\neg$-separated module~$M$ in the effective
topos, the canonical morphism~$M \to \Delta(\Gamma(M))$ is a monomorphism; the
set~$\Gamma(M)$ is flabby by virtue of being inhabited; and~$\Delta$ preserves
flabby objects by Proposition~\ref{prop:pushforward-of-flabby-objects}.
\end{proof}

@@

\item There appears to be some tension regarding the effective topos:
Proposition~\ref{prop:semienough-flabby-modules} shows that at
least~$\neg\neg$-separated modules in the effective topos always embed into
flabby modules, assuming the law of excluded middle in the metatheory, while
Proposition~\ref{prop:flabby-effective-sets} shows that no nontrivial effective
module is flabby.
\end{enumerate}

>>>>

\usepackage[utf8]{inputenc}
\usepackage{amsthm,mathtools,tikz,float,caption,stmaryrd}
\usetikzlibrary{patterns,matrix,decorations.pathreplacing}
\usepackage[all]{xy}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{hyperref}
\usepackage{xspace}

@@

\newcommand{\xra}[1]{\xrightarrow{#1}}
\newcommand{\XXX}[1]{\textbf{XXX: #1}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Mod}{Mod}
\DeclareMathOperator{\Sh}{Sh}
\DeclareMathOperator{\PSh}{PSh}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\effective}{ef{}fective\xspace}

@@

\begin{abstract}
  We introduce a general notion of \emph{flabby objects} in elementary toposes
  and study their basic properties. In the special case of localic toposes, this
  notion reduces to the common notion of flabby sheaves, yielding a
  site-independent characterization of flabby sheaves. Continuing a line of
  research started by Roswitha Harting, we use flabby objects to
  show that an internal notion of injective objects coincides with the
  corresponding external notion, in stark contrast with the situation for
  projective objects. We show as an application that higher direct images can
  be understood as internal cohomology, and study flabby objects in the
  \effective topos.
\end{abstract}

@@

Toby Kenney stressed that the notion of an injective set should be regarded as an
interesting strengthening of the constructively rather ill-behaved notion of a
nonempty set~\cite{kenney:injective-choice}. For instance, while the statements
``there is a choice function for every set of nonempty sets'' and even ``there
is a choice function for every set of inhabited sets'' are constructive taboos,
the statement ``there is a choice function for every set of injective sets'' is
constructively neutral. Proposition~\ref{prop:set-of-preimages-flabby} demonstrates
that the notion of a flabby set can be regarded as an interesting intermediate
notion: In the situation of Proposition~\ref{prop:set-of-preimages-flabby}, the
set of preimages is not only not empty or inhabited, but even flabby.

@@

\section{Flabby objects}
\label{sect:flabby-objects}

@@

If~$R$ is a ring in an elementary topos~$\E$, a similar definition can be given
for~$R$-modules in~$\E$, referring only to the set respectively the object of
linear maps. The condition for an object to be internally injective can be
rephrased in various ways. The following proposition lists five of these
conditions. The equivalence of the first four is due to
Roswitha Harting~\cite{harting:locally-injective}.

@@

The implication (2)~$\Rightarrow$~(2') employs the
extra left adjoint $p_! : \E/A \to \E$ of $p^* : \E
\to \E/A$~(which maps an object~$(X \to A)$ to~$X$), as in the usual proof that
injective sheaves remain injective when
restricted to smaller open subsets: We have that $p_* \circ [\cdot, p^*(I)]_{\E/A}
\cong [\cdot, I]_\E \circ p_!$, the functor $p_!$ preserves monomorphisms, and one
can check that $p_*$ reflects the property that global elements locally possess
preimages. Details are in~\cite[Thm.~1.1]{harting:locally-injective}.\footnote{Harting formulates
the statement for abelian group objects, and has to assume that~$\E$ contains a
natural numbers object to ensure the existence of an abelian version of~$p_!$.}

@@

Somewhat surprisingly, and in stark contrast with the situation for internally
projective objects (which are defined dually), internal injectivity coincides
with external injectivity for localic toposes. In the special case of sheaves
of abelian groups, this result is due to Roswitha
Harting~\cite[Proposition~2.1]{harting:locally-injective}.

\begin{thm}\label{thm:injectivity-external-internal}
Let~$I$ be an object of an elementary topos~$\E$. If~$I$ is externally
injective, then~$I$ is also internally injective. The converse holds if~$\E$ is
localic and Zorn's lemma is available in the metatheory.
\end{thm}

@@

The analogue of Theorem~\ref{thm:injectivity-external-internal} for modules
holds as well, if~$\E$ is assumed to have a natural numbers object. The proof
carries over word for word, only referencing
Lemma~\ref{lemma:set-of-extensions-flabby}(2) instead of
Lemma~\ref{lemma:set-of-extensions-flabby}(1).
It seems that Roswitha Harting was not aware of this generalization, even though she did show that
injectivity of sheaves of modules over topological spaces is a local
notion~\cite[Remark~5]{harting:remark}, as she
(mistakenly) states in~\cite[page~233]{harting:remark} that ``the notions of
injectivity and internal injectivity do not coincide'' for modules.
% XXX comparison with Harting's proof (Barr's lemma...)

@@

\begin{proof}We established the analogous statements for sets and modules purely
intuitionistically in Section~\ref{sect:flabby-sets}, and the stack semantics
is sound with respect to intuitionistic logic.
\end{proof}

@@

\section{Higher direct images as internal sheaf cohomology}
\label{sect:higher-direct-images}

Let~$X$ be a locale and let~$f : Y \to X$ be an over-locale. By the fundamental
relation between locales and topological spaces, this situation arises
for instance, when given a sober topological space and a topological
space over it, as is often the case in algebraic topology or algebraic
geometry. Let a sheaf~$\O_Y$ of rings on~$Y$ be given. Then the
traditional way to define the \emph{higher direct images} of a sheaf~$E$
of~$\O_Y$-modules is to pick an injective resolution~$0 \to E \to I^\bullet$
and set~$R^n f_*(E) \defeq H^n(f_*(I^\bullet))$.

@@

Usually one would not expect an internal construction which depends on
arbitary choices to yield a globally-defined sheaf on~$X$ -- following the
definition of the stack semantics we only obtain a family of sheaves defined on
members of some open covering of~$X$; but we verify in
Theorem~\ref{thm:higher-direct-images-as-internal-sheaf-cohomology} below that in our
case, it does, and that the resulting sheaf coincides with~$R^n f_*(E)$.

@@

The internal characterization provided by
Theorem~\ref{thm:higher-direct-images-as-internal-sheaf-cohomology} gives, as a
simple application, a logical explanation that higher direct images along
the identity~$\mathrm{id} : X \to X$ vanish: From the internal point of view
of~$\Sh(X)$, the over-locale~$X$ corresponds to the one-point locale, and the
higher cohomology of the one-point locale vanishes.

@@

\section{Flabby objects in the \effective topos}
\label{sect:in-eff}

The notion of flabby objects originates from the notion of flabby shaves and is
therefore closely connected to Grothendieck toposes. Hence it is instructive
to study flabby objects in elementary toposes which are not Grothendieck
toposes, away from their original conceptual home. We begin this study with
establishing the following observations on flabby objects in the \effective
topos. We follow the terminology of Martin Hyland's survey on the \effective
topos~\cite{hyland:effective-topos}.

\begin{prop}\label{prop:flabby-effective-sets}
Let~$X$ be a flabby object in the \effective topos. Let~$f : X \to X$
be a morphism. If~$X$ is \effective, the statement ``$f$ has a fixed point''
holds in the \effective topos.
\end{prop}

\begin{prop}\label{prop:semienough-flabby-modules}
Assuming the law of excluded middle in the metatheory, any~$\neg\neg$-separated
module in the \effective topos embeds into a flabby module.
\end{prop}

@@

The intuitive
reason for why Proposition~\ref{prop:flabby-effective-sets} holds is the
following. Let~$X$ be a flabby object in the \effective topos. Then there is a
procedure which computes for any subterminal~$K \subseteq X$ an element~$x_K$
such that~$K \subseteq \{ x_K \}$. This element might not depend extensionally
on~$K$, but this fine point is not important for this discussion. Let
now~$f : X \to X$ be a morphism. We construct the self-referential subset~$K
\defeq \{ f(x_K) \}$; the formal proof below will indicate how this can be
done. Then~$K \subseteq \{ x_K \}$, so~$f(x_K) = x_K$.

A corollary of Proposition~\ref{prop:flabby-effective-sets} is that the trivial
module is the only flabby module in the \effective topos whose underlying
unstructured object is an \effective set: Given such a flabby module~$M$, let~$v
\in M$ be an arbitrary element. Then the morphism~$x \mapsto v + x$ has a fixed
point; thus~$v + x = x$ for some element~$x$, and hence~$v = 0$.

It is the self-referentiality which makes the proof of
Proposition~\ref{prop:flabby-effective-sets} work, but the blame for paucity
of flabby objects in the \effective topos is to put on the realizers for
statements of the form~``$K = K$'', where~$({=})$ is the nonstandard equality
predicate of the powerobject~$\P(X)$. A procedure witnessing flabbiness has to
compute a reflexivity realizer for a suitable element~$x_K$ from a reflexivity
realizer for a given element~$K$. However, such realizers are not very
informative. Metaphorically speaking, a procedure witnessing flabbiness has to
conjure elements out of thin air.

This problem does not manifest with objects~$X$ which are not \effective sets.
Reflexivity realizers for these objects are themselves not very informative;
a procedure witnessing flabbiness therefore only has to turn one kind of
non-informative realizers into another kind. The flabby modules featuring in
the proof of Proposition~\ref{prop:semienough-flabby-modules} will accordingly
not be \effective sets.

\begin{proof}[Proof of Proposition~\ref{prop:flabby-effective-sets}]
For any Turing machine~$e$, let~$v_e : |X| \to \Sigma$ be the nonstandard
predicate given by
\begin{multline*}
  v_e(x) = \{ m \in \NN \,|\,
  \text{there is an element~$x_0 \in |X|$ such that} \\
  \text{$e$ terminates with an element of~$\llbracket x_0 = x_0 \rrbracket$ and
  $m \in \llbracket x = x_0 \rrbracket$} \}
\end{multline*}
and let~$K_e \in \Sigma^{|X| \times \Sigma}$ be the nonstandard predicate given by
\[ K_e(x,u) = \llbracket (x = x) \wedge (u \leftrightarrow v_e(x)) \rrbracket. \]
One can explicitly construct a realizer~$a_e$ of the statement~``$K_e = K_e$'',
where~$({=})$ is the nonstandard equality predicate of the object~$\P_{\leq1}(X)$
of subterminals of~$X$. This is where the assumption that~$X$ is \effective is
important; without it, we could only verify~``$K_e = K_e$'' where~$({=})$ is
the nonstandard equality predicate of the full powerobject~$\P(X)$.

@@

\begin{proof}[Proof of Proposition~\ref{prop:semienough-flabby-modules}]
Let~$(\Gamma \dashv \Delta) : \Set \to \Eff$ be the inclusion of the
double-negation sheaves. For a~$\neg\neg$-separated module~$M$ in the \effective
topos, the canonical morphism~$M \to \Delta(\Gamma(M))$ is a monomorphism; the
set~$\Gamma(M)$ is flabby by virtue of being inhabited; and~$\Delta$ preserves
flabby objects by Proposition~\ref{prop:pushforward-of-flabby-objects}.
\end{proof}

@@

\item There appears to be some tension regarding the \effective topos:
Proposition~\ref{prop:semienough-flabby-modules} shows that at
least~$\neg\neg$-separated modules in the \effective topos always embed into
flabby modules, assuming the law of excluded middle in the metatheory, while
Proposition~\ref{prop:flabby-effective-sets} shows that no nontrivial \effective
module is flabby.
\end{enumerate}
